\large Neural and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2024-02-02 docs/papers/sto/so.tex
FIXUP draft updates

In cosmological $N$-body simulations, the dynamics of the particles is
governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\delta(\vx)$ and $\varphi(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $- \vnabla \varphi$
follows
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
where for each dimension $i\in\{1,2,3\}$ of the 3D field, we then
introduce SO force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of
the wavevector components and all possible relevant features
$\vvartheta$ summarized in \autoref{tab:feat} with more details
discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value for the evenness, and sort the 3 wavevector
compontents to achieve the permutation equivariance.

Being embedded in the differentiable simulation, the NNs can be trained
using gradient-based optimization methods as in usual deep learning tasks.
The parameters of the neural networks, which enter the force evaluation
in every time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.
For both $f$ and $g$ neural networks, we adopt multilayer perceptron
(MLP, i.e. fully connected) structures.

@@

We generate our target data for training and testing with the \GADGET\
simulation code \citep{GADGET-4}.
We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
\YL{
Each simulation evolves $128^3$ particles from the initial time $a_\ic$
to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
We determine $a_\ic$ for each configuration depending on its cosmology
and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why and how here.)} which then sets the
snapshot spacing $\Delta a = (1 - a_\ic) / 64$.
To sample different alignment between the snapshots and the start/stop
times and avoid possible related numerical errors in \pmwd, we introduce
a fractional random offset $\gamma$ that varies between 0 and 1 for
different simulations.
}
In our \pmwd\ model, we generate the snapshot at an arbitrary given time
using the cubic Hermite interpolation, with more details in
\autoref{sec:snapobs}.
The curvature $\OmegaK$ is related to the separate universe simulation
\citep{LiEtAl2014, WagnerEtAl2015}.
\YZ{We refer to this training dataset of 512 \GADGET\ simulations sampled with
Sobol sequence as \GSDATA.}
On average each \GADGET\ simulation takes roughly 2K CPU hours to run, and the
generation of the whole \GSDATA\ dataset consumes 1M CPU hours.
% to be compare with the speed of pmwd with SO

@@

\subsection{Trained Neural Networks \& Symbolic Expressions}

\begin{figure*}
  \centering
  \includegraphics[width=.48\columnwidth]{net_f_a1_j3091776_e3000.pdf}
  \includegraphics[width=.48\columnwidth]{net_g_a1_j3091776_e3000.pdf}
  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
  functions }
  \label{fig:func}
\end{figure*}

@@

@@

Both $f$ and $g$ networks have $7$ hidden layers with $4\times n_{\rm
input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
input layer, which is [...] and [...] for $f$ and $g$ respectively.
A rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply an exponential function to the output as a regulator
for positive output.

@@

\vspace{1em}
\subsection{Symbolic Regression}

We generate a sample of (input, output) data of the trained neural network to
perform feature importance analysis.
We use Sobol sequence to generate the sample of (parameters in
\autoref{tab:param}, $k$, $a$).

>>>>

In cosmological $N$-body simulations, the dynamics of the particles is
governed by the Poisson equation
%
\begin{equation}
  \nabla^2 \varphi(\vx) = \frac32 \Omegam H_0^2 \delta(\vx),
\end{equation}
%
where $\Omegam$ is the matter fraction today, $H_0$ is the Hubble
constant, and the two fields $\varphi(\vx)$ and $\delta(\vx)$ are the
gravitational potential and matter density fluctuation respectively.
In Fourier space, the gravitational force field $- \vnabla \varphi$
follows
%
\begin{equation}
- i \vk \varphi(\vk) = i \frac32 \Omegam H_0^2 \,\frac{\vk}{k^2} \delta(\vk),
\end{equation}
%
where for each dimension $i\in\{1,2,3\}$ of the 3D field, we then
introduce SO force sharpening through
%
\begin{equation}
\frac{i k_i}{k^2} \to \frac{i k_i}{k^2}
  f(k_i; \vvartheta) g(k_1, k_2, k_3; \vvartheta),
\end{equation}
%
where $f$ and $g$ are NNs, each approximating a nonlinear function of
the wavevector components and all possible relevant features
$\vvartheta$ summarized in \autoref{tab:feat} with more details
discussed in \autoref{sec:features}.
Gravity obeys the $\mathrm{O}(3)$ rotational symmetry, which is however
broken by the PM solver to the O\textsubscript{h} discrete symmetry with
48 elements involving permutations and reversals of the 3 principal axes
of the simulation box.
The PM evaluation of each force component can further break the symmetry
along that axis, leaving the remaining 2 permutationally symmetric.
Therefore, we implement $f$ and $g$ to be even functions of the
wavevector components, and $g$ to be permutation equivariant in all 3
components, so that $f g$ is the most general modification to the force
kernel.
An even and real kernel in configuration space is also even in Fourier
space.
We take the absolute value of the wavevectors for the evenness, and sort
the 3 wavevector compontents to achieve the permutation equivariance.

@@

Our target data samples for training and testing are generated with the
\GADGET\ simulation code \citep{GADGET-4}.
We run 512 \GADGET\ simulations that cover a wide range of
configurations and cosmological parameters, as listed in
\autoref{tab:param}.
\YL{
Each simulation evolves $128^3$ particles from the initial time $a_\ic$
to the final time $a_\f=1$, saving 64 snapshots linearly spaced in scale
factor from $a_\ic + \gamma \Delta a$ to $1 - (1 - \gamma) \Delta a$.
We determine $a_\ic$ for each configuration depending on its cosmology
and mean particle spacing \citep{MichauxEtAl2021}, \YL{(YL elaborate why and how here.)} which then sets the
snapshot spacing $\Delta a = (1 - a_\ic) / 64$.
To sample different alignment between the snapshots and the start/stop
times and avoid possible related numerical errors in \pmwd, we introduce
a fractional random offset $\gamma$ that varies between 0 and 1 for
different simulations.
}
In our \pmwd\ model, we generate the snapshot at an arbitrary given time
using the cubic Hermite interpolation, with more details in
\autoref{sec:snapobs}.
The curvature $\OmegaK$ is related to the separate universe simulation
\citep{LiEtAl2014, WagnerEtAl2015}.
\YZ{We refer to this training dataset of 512 \GADGET\ simulations sampled with
Sobol sequence as \GSDATA.}
On average each \GADGET\ simulation takes roughly 2K CPU hours to run, and the
generation of the whole \GSDATA\ dataset consumes 1M CPU hours.
% to be compare with the speed of pmwd with SO

@@

\subsection{Spatial Optimization Neural Networks}

\begin{figure*}
  \centering
  \includegraphics[width=.48\columnwidth]{net_f_a1_j3091776_e3000.pdf}
  \includegraphics[width=.48\columnwidth]{net_g_a1_j3091776_e3000.pdf}
  \caption{$f(k_i; \vvartheta)$ and $g(k_1, k_2, k_3; \vvartheta)$
  functions. The solid line and shaded region indicate the median and
  the [16, 84] percentiles of the 64 Sobol simulation setups. $a=1$,
  mesh shape=1}
  \label{fig:func}
\end{figure*}

For both $f$ and $g$ NNs, we adopt multilayer perceptron (MLP, i.e.
fully connected) structures.
Both $f$ and $g$ networks have $7$ hidden layers with $4\times n_{\rm
input}$ nodes for each layer, where $n_{\rm input}$ is the size of the
input layer, which is [...] and [...] for $f$ and $g$ respectively.
A rectified linear unit (ReLU) activation function is used for hidden
layers.
We further apply an exponential function to the output as a regulator
for positive output.
Being embedded in the differentiable simulation library \pmwd, the NNs
can be trained using gradient-based optimization methods as in usual
deep learning tasks.
The parameters of the NNs, which enter the force evaluation in every
time integration step, can be treated just the same as other
cosmological parameters.
The objective gradients are accumulated, based on which the parameters
are then optimized.

@@

\subsection{Analytic Expressions with Symbolic Regression}

We fit the trained NNs with symbolic regression.

@@

@@

\vspace{1em}
\subsection{Symbolic Regression Fitting of the Neural Networks}

We generate samples of (input, output) data of the trained neural
networks to perform symbolic regression.
We use Sobol sequence to generate the sample of (parameters in
\autoref{tab:param}, $k$, $a$).
